---
title: "Chapter 5"
output: html_document
---

```{r setup, include=FALSE}
library(readr)
library(knitr)
FIREDAM <- read_csv("/home/surya/Downloads/FIREDAM.CSV")
```
#Simple Linear Regression
For this Chapter, our goal is to determine a simple *straight line model* relating a populating mean to another variable.

##Probabilistic Models
Overall, we would want a **deterministic model**, where we can perfectly predict $y$ in terms of $x$.  In general, we can't be exact, so we construct a **probabilistic model**, where 

<center>
$y=$ Deterministic + Random error
</center>

We use **regression modeling**, where we have

<center>
$y=\beta_0+\beta_1 x +\epsilon$
</center>
where

  1. $y$ is the **Dependent** or **Response varable**
  2. $x$ is the **Independent** or **Predictor variable**
  3. $\beta_0 +\beta_1 x$ Deterministic component
  4. $\epsilon$ is the Random error (want to be close to zero)
  5. $\beta_0$ is the **y-intercept**
  6. $\beta_1$ is the **slope of the line**
  
  
##Fitting the Model: The Least Squares Approach

####Example 1
Suppose we have 5 subjects, and we keep track of the percentage of a certain drug in the blodstream and the length of time it takes to react to a stimulus.

```{r}
drug.percent<-c(1,2,3,4,5)
reaction.time<-c(1,1,2,2,4)
cbind(drug.percent,reaction.time)
```

###Scatterplot

```{r}
plot(x = drug.percent, y = reaction.time)
plot(reaction.time~drug.percent)
```

###Correlation
Since as the drug percentage increases, we have an increase in the reaction time, we expect to see a positive correlation, meaning the reaction time is positively correlated to the drug percentage.

```{r}
cor(drug.percent, reaction.time)
```

One the other hand, we could consider the correlation of drug percentage related to the reaction time (which yields the same answer of course!).
```{r}
cor(reaction.time, drug.percent)
```

###Building the Linear Relationship (Regression)


Consider the model
<center>
$\tilde{y} = -1+x$
</center>

Why is this a good model for our data?

<center>
$x$ | $y$ | $\tilde{y} = -1+x$ | $(y-\tilde{y})$ | $(y-\tilde{y})^2$
---|---|---|---|---
1|1|0|(1-0)=1| 1
2|1|1|(1-1)=0|0
3|2|2|(2-2)=0|0
4|2|3|(2-3)=-1|1
5|4|4|(4-4)=0|0
||||Sum of errors = 0| Sum of squared errors (SSE) = 2
</center>

```{r}
plot(reaction.time~drug.percent,xlim=c(0,5),ylim=c(-1,5))
abline(a=-1,b=1)
```


This model is using the **method of least squares**, which calculates the **least squares prediction equation**

###Properties of $\hat{y}=\hat{\beta}_0+\hat{\beta}_1 x$

  1. The sum of the errors equals 0
  2. The sum of the squared errors (SSE) is smaller than any other straight line model
  
###Formulas fo the Least Squares Estimates

  1. *Slope:* $\hat{\beta}_1 = \dfrac{SS_{xy}}{SS_{xx}}$
  2. *y-intercept:* $\hat{\beta}_0 = \stackrel{-}{y}-\hat{\beta}_1\stackrel{-}{x}$
  3. $SS_{xy} = \sum x_iy_i - \dfrac{\sum x_i\sum y_i}{n}$
  4. $SS_{xx} = \sum x_i^2-\dfrac{(\sum x_i)^2}{n}$
  
```{r}
model1<-lm(reaction.time~drug.percent)
model1
```

$x$ | $y$ | $\tilde{y} = -.1+.7x$ | $(y-\tilde{y})$ | $(y-\tilde{y})^2$
---|---|---|---|---
1|1|.6|(1-0)=.4| .16
2|1|1.3|(1-1)=-.3|.09
3|2|2.0|(2-2)=0|0
4|2|2.7|(2-3)=-.7|.49
5|4|3.4|(4-4)=.6|.36
||||Sum of errors = 0| Sum of squared errors (SSE) = 1.10
</center>

```{r}
sum(model1$residuals^2)
anova(model1)
```

```{r}
plot(reaction.time~drug.percent,xlim=c(0,5),ylim=c(-1,5))
abline(model1)
```

If we wanted to use our model to test future values, we use the *predict* function as follows.

```{r}
predict.values<-data.frame(drug.percent = c(8,9,9.5,10))
predict(model1,newdata = predict.values)
```

##Model Assumptions
We have several assumptions for our probability distribution.  

  1. The mean of the probability distribution of $\epsilon$ is 0.
  2. The variance of the probability distribution of $\epsilon$ is constant for all settings of $x$.
  3. The probility distribution of $\epsilon$ is normal
  4. The values of $\epsilon$ associated with any two observed $y$ value are independent.
  
###Estimation of $\sigma^2$ for a (First-Order) Straight-Line Model
We want to find $s$, which will be our **(residual) estimated standard error of the regression model**.  Note that the degrees of freedom for error is $n-2$.

<center>
$SSE = SS_{yy}-\beta_1SS_{xy}$

$s^2=\dfrac{SSE}{n-2}$

$s = \sqrt{\dfrac{SSE}{n-2}}$
</center>

We expect most, about 95%, of the observed y values to lie within 2s of their respective least squares predicted values, $\tilde{y}$.

```{r}
summary(model1)
```
So, s = .61.


##Coefficeints of correlation and Determination
We want to describe the relationship between $x$ and $y$, which is a **bivariate relationship**.  Basically, we want to test the **correlation** between the two variables.  Here, we are going to use correlation to measure the linear relationship between $x$ and $y$.  Our two variables of interest are

  * $r$ - coefficient of correlation
  * $r^2$ - coefficient of determination
  
with formulas


  1. $r=\dfrac{SS_{xy}}{\sqrt{SS_{xx}SS_{xy}}}$
  2. $r^2=\dfrac{SS_{yy}-SSE}{SS_{yy}} = 1-\dfrac{SSE}{SS_{yy}} = 1-\dfrac{SSE}{TSS} = \dfrac{SSR}{TSS}$
  3. TSS = $\sum(y-\stackrel{-}{y})^2$
  4. SRR = TSS - SSE

Notice that $-1<r<1$ and $0<r^2<1$.  When $r$ is near 0, no there is no relationship between $x$ and $y$.  When $r$ is near $1$, we have a positive correlation, and being near $-1$ gives a negative correlation.  When $r^2=a$, we have that a% of the sample variation of $y$ can be explained by using $x$ to predict $y$ (in our straight-line model).

By hand, we can calculate $r^2$ as follows:

```{r}
SSE<-sum(summary(model1)$residuals^2)
TSS<-sum( ((reaction.time) - mean(reaction.time))^2)
SSR<-TSS-SSE
SSR/TSS
```

SSE|TSS|SSR
---|---|---
1.1|6|4.9

Or we can use the ouput from the *summary* of our linear model.

```{r}
summary(model1)$r.squared
```
Since $r=.90$ is positive and near 1, we have that the reaction time tends to increase as the drug percent increases.  Also, since $r=.82$, 82% of the sample variation in reaction time ($y$) can be explained by using the percent of the drug ($x$) in a straight line model.


####Example 2
We want to find a relationship between the distance between a burning hourse and the nearest fire station and the amount of fire damage.  We'll set $x$ to be the distance from the fire station and $y$ to be the amount of damage (using FIREDAM data set).  

  1. Find the linear model and estimate your approximation.
  2. Find $r$ and $r^2$ and interpret.

```{r}
fire.model<-lm(DAMAGE~DISTANCE,FIREDAM)
summary(fire.model)
plot(DAMAGE~DISTANCE,FIREDAM)
abline(fire.model)
text(x=5,y=20,label = "y = 10.27 + 4.91x ")
```
```{r}
sum(fire.model$residuals^2)
summary(fire.model)$sigma
cor(FIREDAM$DISTANCE,FIREDAM$DAMAGE)
summary(fire.model)$r.squared
```

###Summary
In general, the best-fit line for the damage of the fire given the nearest firestation is $y=10.27+4.91x$.  The standard error is 2.31, meaning that all of our data points fall within $\pm 4.62$ from our line. Our correlation and determination factors are 96% and 92%.  This mean that 96% of the damage is correlated to the distance of the closest fire station and 92% of the damage can be determined by that same distance.  