---
title: "Chapter 3"
output: html_document
---

```{r setup, include=FALSE}
library(readr)
library(knitr)
library(plyr)
library(qcc)
APHASIA <- read_csv("C:/Users/mgreer/Dropbox/Mark/UNA/Courses/MA345 - Applied Statistics/Datasets/CSV/Examples/APHASIA.CSV")
EPAGAS <- read_csv("C:/Users/mgreer/Dropbox/Mark/UNA/Courses/MA345 - Applied Statistics/Datasets/CSV/Examples/EPAGAS.CSV")
Finch_By_Year_Data <- read_csv("C:/Users/mgreer/Dropbox/Mark/UNA/Courses/MA345 - Applied Statistics/Datasets/Hawkes DataSets/Finch_By_Year_Data.csv")
```

## Frequency Distributions

Since we can't do "normal" mathematical operations with qualitative data, we focus on classifying the data and visually organizing the data.  We have

  * **classes** categories of the experimental units
  * **frequency** the number of occurrences in each class
  * **relative frequency** the frequency divided by the total size of the sample
  * **cumulative frequency** the sum of the frequency of a particular class and all preceding classes
  * **cumulative relative frequency** the proportion of observations in a particular class and all preceding classes
  


#### Example 1 
Using the APHASIA file, will classify first classify the data.  The *count* function comes from the **plyr** library.

```{r}
c1<-count(APHASIA)
c1
data<-c1$freq
```

Notice that the count function gives the classes and frequencies.  To get the relative , cumulative, and cumulative relative frequency, we compute:

```{r}
rel.freq = data/sum(data)
rel.freq = round(rel.freq,2)
cum.freq = cumsum(data)
cum.rel.freq = cumsum(rel.freq)
rel.freq
cum.freq
cum.rel.freq
```

Now we'll constuct a table using the *kable* function in the **knitr** library.

```{r}
mytable<-data.frame("Names" = c1$TYPE, "Frequency" = data)
kable(mytable)
```

To add the other columns (for example), use the *cbind* function.

```{r}
mytable<-cbind(mytable,"Cumulative Frequency" = cum.freq, "Relative Frequency" = rel.freq, "Cumulative Relative Frequency" = cum.rel.freq)
kable(mytable)
```

## Displaying Qualitative Data Graphically (Bar Plots, Pie Graphs, and Pareto Diagrams)

Below are the commands to constuction different types of charts for our data.  Please utilize the manual, internet, etc. to find ways to change and improve these to your liking.

```{r}
barplot(data, names.arg = c1$TYPE)
pie(data,labels = paste(c1$TYPE,rel.freq *100,"%"))
```

For a pareto diagram, we'll use the *pareto.chart* function from the **qcc** library.

```{r}
names(data)<-c1$TYPE
pareto.chart(data)
```
## Histograms and Other Graphical Displays of Quantitative Data
We can also visual quantitative data.  Frequency tables are made the same was as before, except now we must make classes as intervals for our data (not *named* classes as with quantitative).

#### Example 2
Using the EPAGAS file

```{r}
gas.mpg<-EPAGAS$MPG
breaks<-seq(30,45,1)
Classes<-cut(gas.mpg,breaks,right=F)
mytable2<-data.frame(table(Classes))
kable(mytable2)
```

We can add other colums as before.  Now we show several functions for visualizing quantitative data.


```{r}
dotchart(table(Classes))
stem(gas.mpg,scale=2)
hist(gas.mpg,breaks=14)
hist(gas.mpg,breaks=14,prob=T)
lines(density(gas.mpg))
```

#### Example 3
Finally we have a time series graph.

```{r}
ts_example<-ts(Finch_By_Year_Data$`Beak length`[Finch_By_Year_Data$Species == "fortis"], frequency = 1, start = c(1973,1))
plot.ts(ts_example)
```