---
title: "Chapter 4"
output: html_document
---

```{r setup, include=FALSE}
library(readr)
library(knitr)
library(plyr)
library(qcc)
library(zoo)
EPAGAS <- read_csv("C:/Users/mgreer/Dropbox/Mark/UNA/Courses/MA345 - Applied Statistics/Datasets/CSV/Examples/EPAGAS.CSV")
Finch_By_Year_Data <- read_csv("C:/Users/mgreer/Dropbox/Mark/UNA/Courses/MA345 - Applied Statistics/Datasets/Hawkes DataSets/Finch_By_Year_Data.csv")
```

## Measure of Location
Given a quantitative data set, we want a good candidate to represent the data.  Below are several examples that are normally choosen

  * **mean** average data $\dfrac{\sum x_i}{n}$
    + $\stackrel{-}{x}$ sample mean
    + $\mu$ population mean
  * **weighted mean** $\dfrac{\sum w_ix_i}{\sum w_i}$
  * **trimmed mean** modification to the mean by ignoring a certain percentage of the highest and lowest data values.
  * **median** the middle value
  * **mode** the value that occurs the most (might be multiple)
  

```{r include=FALSE}
gas.mpg<-EPAGAS$MPG
```

```{r}
mean(gas.mpg)
mean(gas.mpg, trim = .2)
median(gas.mpg)
```

For the weighted mean, consider a GPA calculation (4.0 scale) with letter grades B, A, C, A and credits $\{3, 4, 5, 3\}$ respectively.

```{r}
grades<-c(3,4,2,4)
grades.weight<-c(3,4,5,3)
weighted.mean(grades,grades.weight)
```

### Relationship between the Mean, Median, and Mode
Usually the shape of the data determines how the mean, median, and mode are related.  Data can be skewed as

  * **right or positive skewed** mean falls to the right of the median and mode falls to the left
  * **left or negative skewed** mean false to the left of the median and mode falls to the right
  * **symmetric skewed** mean, median, and mode are all the same

### Time Series 
For time series data, often looking at consecutative data points is helpful rather than an overall calculation

  * **moving average** mean over a certain time frame (using the library *zoo*).
  
```{r}
beak.length <- Finch_By_Year_Data$`Beak length`[Finch_By_Year_Data$Species == "fortis"]
mean(beak.length)
rollmean(beak.length, 2)
```

```{r}
ts_example<-ts(Finch_By_Year_Data$`Beak length`[Finch_By_Year_Data$Species == "fortis"], frequency = 1, start = c(1973,1))
plot.ts(ts_example)
plot(beak.length)
lines(rollmean(beak.length,2), col = c("red"))
lines(rollmean(beak.length,3), col = c("blue"))
```

## Measure of Dispersion
Given a quantitative data set, we want to know the spread of data.  Below are several examples that are normally choosen.

  * **range** the difference between the largest and smallest data values.
  * **Mean Absolute Deviation (MAD)** $\dfrac{\sum |x_i - \stackrel{-}{x}|}{n}$
  * **variance** average squared distance from the mean $\dfrac{\sum(x-\stackrel{-}{x})^{2}}{n-1}$
    + $s^2$ sample variance
    + $\sigma^2$ population variance (divide by $N$ not $N-1$)
  * **standard deviation** square root of the variance $\sqrt{\sigma^{2}}$ or $\sqrt{s^2}$
    + $s$ sample standard deviation
    + $\sigma$ population standard deviation
  
```{r}
range(gas.mpg)
range(gas.mpg)[2]-range(gas.mpg)[1]
mad(gas.mpg)
var(gas.mpg)
sd(gas.mpg)
```

### Emperical Rule
Using the standard deviation, we can easily estimate where the data falls.  If the data is *symmetic*

  * $68\%$ of the data falls within 1 standard deviation away from the mean
  * $95\%$ of the data falls within 2 standard deviation away from the mean
  * $99.7\%$ of the data falls within 3 standard deviation away from the mean
  
```{r echo=FALSE}
cord.x1 <- c(-1,seq(-1,1,0.01),1) 
cord.y1 <- c(0,dnorm(seq(-1,1,0.01)),0) 

cord.x2a <- c(-2,seq(-2,-1,0.01),-1) 
cord.y2a <- c(0,dnorm(seq(-2,-1,0.01)),0) 
cord.x2b <- c(1,seq(1,2,0.01),2) 
cord.y2b <- c(0,dnorm(seq(1,2,0.01)),0) 

cord.x3a <- c(-3,seq(-3,-2,0.01),-2) 
cord.y3a <- c(0,dnorm(seq(-3,-2,0.01)),0) 
cord.x3b <- c(2,seq(2,3,0.01),3) 
cord.y3b <- c(0,dnorm(seq(2,3,0.01)),0) 

curve(dnorm(x,0,1), xlim=c(-3,3), main='Standard Normal') 
 
polygon(cord.x1,cord.y1,col='green')
polygon(cord.x2a,cord.y2a,col='red')
polygon(cord.x2b,cord.y2b,col='red')

polygon(cord.x3a,cord.y3a,col='blue')
polygon(cord.x3b,cord.y3b,col='blue')
```
  
### Chebyshev's Theorem

If the data is skewed, or we don't know if it's symmetric, then we can use **Chebyshev's Rule/Theorem**

  * at least $1-\dfrac{1}{k^2}$ of the data falls within $k$ standard deviations away from the mean ($k>1$)
  
### Coefficient of Variation
It is often useful to compare variation between two or more data sets.  The **coefficient of variation** acheives this in a standard way, by dividing by the mean.  The main advantage to the $CV$ is that it is unitless, so it can be compared to each other that others measures can't be (*e.g.* standard deviation, etc.)

  * **coefficient of variation** Population $CV = \left( \frac{\sigma}{\mu}\cdot 100\right)\%$ or Sample $\left( \frac{s}{\stackrel{-}{x}}\cdot 100\right)\%$
  
```{r}
sd(gas.mpg)/mean(gas.mpg)*100
```
  
## Percentiles and Quartiles
The **p^th^ percentile** is the value such that $p\%$ of the measurements fall below that value and $(1-p)\%$ falls above that value.  **Quartiles** are simply the $25\%,50\%,75\%$ percentiles ($Q_1, Q_2, Q_3$).

```{r}
quantile(gas.mpg,c(.05,.35,.76,.89))
quantile(gas.mpg)
```

The difference between $Q_3$ and $Q_1$ (i.e. where $50\%$ of the data lives) quartiles is called the $IQR$.

```{r}
IQR<-quantile(gas.mpg)[4]-quantile(gas.mpg)[2]
```

##Box and Whisker Plots
Box and Whisker plots are also used to find outliers.  It is created by using the median of a data set, with the box being made with the $25\%$ and $75\%$ percentiles.  The whiskers are made by subtracting $1.5*IQR$ to the $25\%$ percentile and adding $1.5*IQR$ to the $75\%$ percentile.  Values falling above and below these whiskers are usually seen to be outliers.

```{r}
boxplot(gas.mpg,horizontal = T)
summary(fivenum(gas.mpg))
```

```{r}
IQR<-quantile(gas.mpg)[4]-quantile(gas.mpg)[2]
lbs<-c(quantile(gas.mpg)[2]-1.5*IQR,quantile(gas.mpg)[2],median(gas.mpg),quantile(gas.mpg)[4],quantile(gas.mpg)[4]+1.5*IQR)
boxplot(gas.mpg,horizontal = T)
text(lbs,labels=lbs,y = 1.25)
```

The following gives the information above as well as a nice way to view the outliers.

```{r}
boxplot.stats(gas.mpg)
boxplot(gas.mpg,horizontal = T)
lbs<-boxplot.stats(gas.mpg)$stats
text(lbs,labels=lbs, y = 1.25)
```

Again, you can see above that we have two suspected outliers.

## Z-Scores
Comparing data sets or individual data can be done using **z-scores**.  Z-scores scale the data set so that the spread is the same, but the mean and standard deviation are set to 0 and 1 respectively.  In this setting, data sets with different means and standard deviations can be compared objectively.  The formula is

  * **z-score of a sample** $\dfrac{x-\stackrel{-}{x}}{s}$
  * **z-score of a population** $\dfrac{x-\mu}{\sigma}$
  
```{r}
scale(gas.mpg)
min(scale(gas.mpg))
max(scale(gas.mpg))
mean(scale(gas.mpg))
sd(scale(gas.mpg))
```

Z-scores can also tell use about outliers.  Usually, if the z-score is close/below -3 or close/above 3, then we usually assume that value is an outlier.  You can see we have two values that we would usually assume are outliers.

## Proportions and Percentages
A **proportion** is the measure that a fraction of a group possesses some characteristic.

  * $X =$ the number that possess the characteristic
  * $N =$ the number in the population (or sample)
  * $p = \frac{X}{N}$ the population proportion
  * $\hat{p} = \frac{X}{n}$ the sample proportion
